Fast Sorting on Modern Computers

### 2005-05-30 paper/docs/heap.tex
added blurb and sample code for the software predictors

Heapsort is an in-place sort which works in $O(NlogN)$ time in the average and
worst case. It begins by creating a structure known as a \n{heap}, which stores
its largest key at the top, then removing keys from the top and placing them in
their final position in the array.

@@

The first method involves \n{sifting}\footnote{Sifting up and down is also
referred to as fixing up and down, and algorithms that do this are called
\n{fix-up} or \n{sift-up}, and \n{fix-down} or \n{sift-down}} keys up the heap.
Assuming that a heap already exists, adding a key and re-heapifying will
maintain the heap property. This can be done by adding the key at the end of
the heap, then moving the key up the tree, swapping it with smaller parents
until it is in place. An unsorted array can be sorted by continually heapifying
larger lists, thereby implicitly adding one key every time to a sorted heap of
initial size one. This technique was invented by Williams, and LaMarca refers to
it as the \n{Repeated-Adds} algorithm.

@@

The second step is to destroy the heap, leaving a sorted array.

@@

LaMarca overcomes the problem by inserting a maximal key at the end of a heap if
there's an even number of nodes in the array. This gives a second child to the
last element of the heap. This is an effective technique for building the heap,
as the number of keys is static. When destroying the heap, however, the
maximal key must be added to every second operation. This can be avoided
relatively simply with no extra instructions. If the last key in the heap is
left in the heap while its being sifted down the heap, it will provide a
sentinel for the sift down operation. The key from the top of the heap is then
put into place at the end. This technique reduces instruction count by
approximately 10\%.

@@

LaMarca notes that the Repeated-Adds algorithm has a much lower number of cache
misses than Floyd's. The reason for this is simple. Consider the number of keys 
touched when a new key is added to the heap. In the case of Floyd's method, the
keys being heapified are not related to the last keys that were heapified. Reuse
only occurs upon reaching a much higher level of the heap. For large heaps, these
keys may be swapped out of the cache by this time. As a result, Floyd's method
has poor temporal locality.

The Repeated-Adds algorithm, however, displays excellent temporal locality. A
key has a 50\% chance of having same parent, a 75\% of the same grandparent and
so on. The number of instructions, however, increases significantly, and the
sort becomes out-of-place.

The cost of building the heap is completely dominated by the cost of
destroying it. Less than 10\% of heapsort's time is spent building the heap.
Therefore optimisations must be applied when destroying the heap.

@@

@@

In the graph, the base heapsort's results flat-line for a step longer than the
other heapsorts' results. This is due to being in-place, since twice as many
keys fit in the cache without a corresponding increase in misses. For the same
reason, the base heapsorts have half as many level 2 cache misses of the later,
out-of-place sorts (See Figure \ref{heap level 2 misses}).

@@

The difference between different types of predictors is also visible in the
results, shown in Figure \ref{heap branch misses}. The largest table size is
shown for the two-level predictor. The smallest table size is shown for the
bimodal predictor, as all the bimodal table sizes show identical results.  In
the case of the 2- and 8-heaps, the bimodal predictor works better than the
two-level adaptive predictor. The warm-up time is visible in the base heapsort,
though it's amortised for larger data sets. There is no difference between
predictors for the 4-heap.

Figure \ref{heap cycles} shows the actual time taken for heap sort on a Pentium
4 1.8GHz machine, which has an 8-way associative 256KB level 2 cache with a 64 byte
cache line, a 4-way associative 8KB level 1 data cache, and a maximum branch
misprediction delay of 20 cycles.\footnote{This is determined from the Pentium
4's 20 stage pipeline. This number is not exact, and is probably slightly lower,
with the possibility of occasionally rising higher (in the case of an
instruction cache miss, which is exceptionally unlikely for small routines like
sorts).} Its static branch prediction strategy is backward taken and forward
not taken; its dynamic strategy is unspecified: though \cite{Intel248966-010}
mentions a branch history so it is likely a two-level adaptive predictor.

The figure shows that the base heapsort performs best while the data set fits in
the cache. After that point, the 8-heap performs best, and its cycle count increases
with a smaller slope. This shows the importance of LaMarca's cache-conscious
design even eight years after his thesis. It also shows the importance of reduced
branch mispredictions: the base heapsort has more instructions and level 1 cache
misses, but still outperforms the 4-heap, before level 2 cache misses take over.
However, this is hardly conclusive: the lower level 2 cache misses may account
for this.

@@

Finally, though this is likely to be related the previous differences, base
heapsort performs slightly better for small sets than the cache heapsorts do in
the results presented here. LaMarca's results indicate that the cache heapsort
has better results the entire way through the cache. This may relate to the
associativity: LaMarca's tests were performed on a DEC AlphaStation 250 with a
direct mapped cache. The Pentium 4's cache is 8-way associative, and so the
unoptimised algorithm would not suffer as much.

\section{Predictor results}

@@

The most interesting thing is the relation between the predictivity of the
comparisons in the destruction of the 2-, 4- and 8-heaps, which are shown in
Figure \ref{Branch prediction performance for base and cache heapsorts}.
Overall, the 4-heap is half as high as the 2-heap, and twice as high as the
8-heap, but has to make more comparisons. However, there are three comparisons
in the 4-heap and seven comparisons in the 8-heap. In total, there are 50\% more
comparisons in the 4-heap, and 150\% more in the 8-heap.

@@

LaMarca later published heapsort code in \cite{LaMarcaHeap96}. The ACM
Journal of Experimental Algorithms requires that source code be published, and
comparing simulated results from there to the sorts created here might have
interesting results.

>>>>

Heapsort is an in-place sort which works in $O(NlogN)$ time in the average and
worst case. It begins by creating a structure known as a \n{heap}, a binary tree
in which a parent is always larger its child, and the largest key is at the top
of the heap. It then repeatedly removes the largest key from the top and places
it in its final position in the array, fixing the heap as it goes.

@@

The first method involves \n{sifting}\footnote{Sifting up and down is also
referred to as fixing up and down, and algorithms that do this are called
\n{fix-up} or \n{sift-up}, and \n{fix-down} or \n{sift-down}.} keys up the heap.
Assuming that a heap already exists, adding a key and re-heapifying will
maintain the heap property. This can be done by adding the key at the end of
the heap, then moving the key up the tree, swapping it with smaller parents
until it is in place. An unsorted array can be sorted by continually heapifying
larger lists, thereby implicitly adding one key every time to a sorted heap of
initial size one. This technique was invented by Williams, and LaMarca refers to
it as the \n{Repeated-Adds} algorithm.

@@

The second step, described next, is to destroy the heap, leaving a sorted array.

@@

LaMarca overcomes the problem by inserting a maximal key at the end of a heap if
there's an even number of nodes in the array. This gives a second child to the
last element of the heap. This is an effective technique for building the heap,
as the number of keys is static. When destroying the heap, however, the
maximal key must be added to every second operation. This can be avoided
relatively simply with no extra instructions. If the last key in the heap is
left in place aswell as being sifted down the heap, it will provide a
sentinel for the sift-down operation. The key from the top of the heap is then
put into place at the end. This technique reduces instruction count by
approximately 10\%.

@@

LaMarca notes that the Repeated-Adds algorithm has a much lower number of cache
misses than Floyd's. The reason for this is simple. Consider the number of keys 
touched when a new key is added to the heap. In the case of Floyd's method, the
key being heapified is not related to the last key that was heapified. Reuse
only occurs upon reaching a much higher level of the heap. For large heaps, these
keys may be swapped out of the cache by this time. As a result, Floyd's method
has poor temporal locality.

The Repeated-Adds algorithm, however, displays excellent temporal locality. A
key has a 50\% chance of having the same parent, a 75\% of the same grandparent
and so on. The number of instructions, however, increases significantly, and the
sort becomes out-of-place. For this reason, the Repeated-Adds algorithm is used
for cache heapsort.

The cost of building the heap is completely dominated by the cost of
destroying it. Less than 10\% of heapsort's time is spent building the heap.
Therefore optimizations must be applied when destroying the heap.

@@

%TODO this isnt true is it? placing the nsentinel is cheaper than a bounds check
%(a write to a variable thats going to be written to in a second vs a comparison?)
%and there arent an extra logN instructions, theres logN times as many

@@

In the graph, the base heapsort's results flat-line for a step longer than the
other heapsorts' results. This is due to being in-place, since twice as many
keys fit in the cache without a corresponding increase in misses. For the same
reason, the base heapsorts have approximately half as many level 2 cache misses
as the out-of-place sorts (See Figure \ref{heap level 2 misses}).

@@

The difference between different types of predictors is also visible in the
results, shown in Figure \ref{heap branch misses}. The largest table size is
shown for the two-level predictor. The smallest table size is shown for the
bimodal predictor, as all the bimodal table sizes show identical results.  In
the case of the 2- and 8-heaps, the bimodal predictor works better than the
two-level adaptive predictor. The warm-up time of the table-driven predictors is
visible in the base heapsort results, though it's amortised for larger data
sets.

Figure \ref{heap cycles} shows the actual time taken for heap sort on a Pentium
4 1.8GHz machine, which has an 8-way associative 256KB level 2 cache with a 64 byte
cache line, a 4-way associative 8KB level 1 data cache, and a maximum branch
misprediction delay of 20 cycles\footnote{This is determined from the Pentium
4's 20 stage pipeline. This number is not exact, and is probably slightly lower,
with the possibility of occasionally rising higher (in the case of an
instruction cache miss, which is exceptionally unlikely for small routines like
sorts).}. Its static branch prediction strategy is backward taken and forward
not taken; its dynamic strategy is unspecified: though \cite{Intel248966-010}
mentions a branch history so it is likely a two-level adaptive predictor.

The figure shows that the base heapsort performs best when the data set fits in
the cache. After that point, the 8-heap performs best, and its cycle count increases
with a smaller slope. This shows the importance of LaMarca's cache-conscious
design even nine years after his thesis. It also shows the importance of reduced
branch mispredictions: the base heapsort has more instructions and level 1 cache
misses, but still outperforms the 4-heap, until level 2 cache misses take over.
However, this is hardly conclusive: the lower level 2 cache misses may account
for this.

@@

Finally, though this is likely to be related the previous differences, base
heapsort performs slightly better for small sets than the cache heapsorts do in
the results presented here. LaMarca's results indicate that the cache heapsort
has better results the entire way through the cache. This may relate to the
associativity: LaMarca's tests were performed on a DEC AlphaStation 250 with a
direct mapped cache. The Pentium 4's cache is 8-way associative, and so the
unoptimized algorithm would not suffer as much.

\section{Predictor Results}

@@

The most interesting thing is the relation between the predictablility of the
comparisons in the destruction of the 2-, 4- and 8-heaps, which are shown in
Figure \ref{Branch prediction performance for base and cache heapsorts}.
Overall, the 4-heap is half as high as the 2-heap, and twice as high as the
8-heap, but has to make more comparisons. However, there are three comparisons
in the 4-heap and seven comparisons in the 8-heap. In total, there are 50\% more
comparisons in the 4-heap, and 150\% more in the 8-heap.

@@

LaMarca later published heapsort code in \cite{LaMarcaHeap96}. The ACM
Journal of Experimental Algorithms requires that source code be published, and
comparing simulated results from there to the sorts created here might have
interesting results and would explain the discreptancies highlighted in this
chapter.

### 2005-05-30 paper/docs/method.tex
added blurb and sample code for the software predictors

@@

>>>>

Finally, a software predictor was added in order to pin-point the cause of the
branch mispredictions. This meant altering the code of the algorithms to call
functions which updated simulated branch predictors. Each important branch was
allocated a predictor, which was updated independantly of the other predictors.
This allowed the performance to be analysed down to the individual branch, and
blame apportioned appropriately.

@@

\subsection{Software Predictor}
The software predictor used was a series of macros and functions which
maintained the state of a branch predictor at a particular point in the program.
Statistics about each branch were collected, and dumped into text files, from
which scripts created graphs to display them. Some sample code is shown in
Figure \ref{Sample code using the software predictor}.

\code{Sample code using the software predictor}{quick.c}
